---
title: "Practical - Introduction to Mathematical Modelling"
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    toc-expand: 2
editor: visual
---

```{r, message=FALSE}
# Load required libraries
library(dplyr)
library(deSolve)
library(ggplot2)
library(gt)
```

# Estimating $\mathcal{R}_0$ from the final size of an epidemic

The final proportion of a population infected ($r_\infty$) and the basic reproduction number ($\mathcal{R}_0$) are related through the following equation:

\begin{equation}

  r_\infty = 1 - \exp(-\mathcal{R}_0 r_\infty)

\end{equation}

To solve for $\mathcal{R}_0$ given a final size we can simply plug in our observed final size and rearrange the equation as shown below:

\begin{equation}

  \mathcal{R}_0 = -\frac{\ln(1 - r_\infty)}{r_\infty}

\end{equation}

## Question 1

Write a function called `my_R0` that takes a value of $r_\infty$ (the final epidemic size) and returns an estimated $\mathcal{R}_0$ based on the above relationship.

```{r}
my_R0 <- function(r_inf)
{
  # Your code here
  # Hint: Use log() for natural logarithm in R
  return()
}
```

## Question 2

```{r}
diseases <- data.frame(
  Disease = c("H1N1 Influenza", "Measles"),
  Proportion_Infected = c(0.24, 0.77),
  Source = c("Lessler et al, 2009", "Panum, 1847")
)

diseases |>
  gt() |>
  cols_label(
    Proportion_Infected = "Proportion Infected") |>
  fmt_number(
    columns = Proportion_Infected,
    decimals = 2)
```

Use `my_R0` to estimate $\mathcal{R}_0$ for each of the two outbreaks above.

```{r}
# Write code here

```

## Question 3

Next, let’s estimate the final size of an epidemic for a given value of $\mathcal{R}_0$. To do this, we need to solve the equation above for $r_\infty$ using an iterative method. Start by writing an objective function that takes a value of $r_\infty$ and $\mathcal{R}_0$ as inputs and returns the absolute difference between the right and left hand sides of the equation above. We will want to minimise this value (i.e. the smallest difference between the right side and left sides of the equation). Try three candidate values for $r_\infty$ of 0.2, 0.4, 0.6, with an $\mathcal{R}_0$ of 1.1 - which is the closest to the optimal?

```{r}
objective_function <- function(r_inf, R0)
{
  # Calculate left hand side of equation
  left_side <-
  # Calculate right hand side of equation
  right_side <-
  # Return absolute difference
  return()
}
r_inf_test_1 <- 0.2
r_inf_test_2 <- 0.4
r_inf_test_3 <- 0.6

# Test your function here

# Find optimal r_inf using optimization
fixed_R0 <- 1.1

# Uncomment the lines below
#wrapper_objective_fn <- function(r_inf) objective_function(r_inf, fixed_R0)
#res <- optimise(wrapper_objective_fn, interval = c(0, 1))
```

## Question 4

For three different pandemic influenza outbreaks, the estimated mean $\mathcal{R}_0$ values were 1.5, 2.0, and 2.5. Estimate the final epidemic size for each outbreak.

```{r}
# Code here

```

# Numerically solving SIR models

We will introduce code to numerically integrate SIR models. We will use the deSolve package.

## Question 5

Create an SIR function that returns dS, dI, and dR for a given set of inputs: par_beta, par_gamma, and par_N (population size). Note that here we are working with the actual numbers of susceptible, infectious, and recovered individuals rather than their proportions.

```{r}
sir_model <- function(time, compartments, parameters) {
  with(as.list(c(compartments, parameters)), {
    # derivatives
    dS <- 
    dI <-  
    dR <-  
    
    list(c(dS, dI, dR))
  })
}
```

## Question 6

We can then use this code to numerically integrate an SIR model with fixed parameters. Use an initial state of `S = 9999`, `I = 1`, `R = 0`, a `gamma of 0.5` and a `beta of 1`. Plot the resulting curves of S, I and R. Simulate for 60 days.

```{r}
# Initial values
#par_N        <- 
#compartments <- 

# Parameters
parameters <- c(
  #par_beta = ,   # transmission rate
  #par_gamma = ,  # recovery rate
  #par_N = par_N # population size
)

# Time points
#stop_time <- 
#times <- seq(0, stop_time, by = 1/8)

# Solve ODE (uncomment lines below)
#out <- ode() |> as.data.frame()
```

## Question 7

Run two more simulations. Change `par_beta` to `0.75` and `1.25`. All other parameters and states should be the same as above. Plot the output from these simulations on the same plot. Describe how these two epidemics differ. Why would a different beta value result in a different epidemic?

```{r}
# Code here
# Hint: Run two additional simulations with different beta values
# Hint: Combine results into one data frame for plotting
```

## Question 8

Estimate the final epidemic size of the three simulations above and compare against those estimated in Question 4.

```{r}
# Code here
```

## Question 9

What's the largest value of `par_beta` for which the introduction of an infectious individuals does not result in an epidemic? Show your answer with a simulation. Simulate for 2000 days.

```{r}
# Code here
```
