---
title: "Practical - Introduction to Mathematical Modelling (Solutions)"
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    toc-expand: 2
editor: visual
---

```{r, message=FALSE}
# Load required libraries
library(dplyr)
library(deSolve)
library(ggplot2)
library(gt)
library(tidyr)
```

# Estimating $\mathcal{R}_0$ from the final size of an epidemic

The final proportion of a population infected ($r_\infty$) and the basic
reproduction number ($\mathcal{R}_0$) are related through the following equation:

\begin{equation}

  r_\infty = 1 - \exp(-\mathcal{R}_0 r_\infty)

\end{equation}

To solve for $\mathcal{R}_0$ given a final size we can simply plug in our observed
final size and rearrange the equation as shown below:

\begin{equation}

  \mathcal{R}_0 = -\frac{\ln(1 - r_\infty)}{r_\infty}

\end{equation}

## Question 1

Write a function called `my_R0` that takes a value of $r_\infty$ (the final 
epidemic size) and returns an estimated $\mathcal{R}_0$ based on the above 
relationship.

```{r}
my_R0 <- function(r_inf)
{
  R0 <- -log(1 - r_inf) / r_inf
  return(R0)
}
```

## Question 2

```{r}
diseases <- data.frame(
  Disease = c("H1N1 Influenza", "Measles"),
  Proportion_Infected = c(0.24, 0.77),
  Source = c("Lessler et al, 2009", "Panum, 1847")
)

diseases |>
  gt() |>
  cols_label(
    Proportion_Infected = "Proportion Infected") |>
  fmt_number(
    columns = Proportion_Infected,
    decimals = 2)
```

Use `my_R0` to estimate $\mathcal{R}_0$ for each of the two outbreaks above.

```{r}
my_R0(diseases$Proportion_Infected[[1]]) |> round(2)
my_R0(diseases$Proportion_Infected[[2]]) |> round(2)
```

## Question 3

Next, let’s estimate the final size of an epidemic for a given value of 
$\mathcal{R}_0$. To do this, we need to solve the equation above for $r_\infty$ 
using an iterative method. Start by writing an objective function that takes a value of 
$r_\infty$ and $\mathcal{R}_0$ as inputs and returns the absolute difference 
between the right and left hand sides of the equation above. We will want to 
minimise this value (i.e. the smallest difference between the right side and 
left sides of the equation). Try three candidate values for $r_\infty$ of 0.2, 
0.4, 0.6, with an $\mathcal{R}_0$ of 1.1 - which is the closest to the optimal?

```{r}
objective_function <- function(r_inf, R0)
{
  left_side <- r_inf
  right_side <- 1 - exp(-R0 * r_inf)
  return(abs(left_side - right_side))
}

R0_test <- 1.1
r_inf_test_1 <- 0.2
r_inf_test_2 <- 0.4
r_inf_test_3 <- 0.6

result_1 <- objective_function(r_inf_test_1, R0_test)
result_2 <- objective_function(r_inf_test_2, R0_test)
result_3 <- objective_function(r_inf_test_3, R0_test)

cat("Test results for R0 = 1.1:\n")
cat("r_inf = 0.2:", result_1, "\n")
cat("r_inf = 0.4:", result_2, "\n")
cat("r_inf = 0.6:", result_3, "\n")

# Find optimal r_inf using optimization
fixed_R0 <- 1.1
wrapper_objective_fn <- function(r_inf) objective_function(r_inf, fixed_R0)
res <- optimize(wrapper_objective_fn, interval = c(0, 1))

cat("\nOptimization result:\n")
cat("Minimum objective value:", res$objective, "\n")
cat("Optimal r_inf:", res$minimum, "\n")
```

## Question 4

For three different pandemic influenza outbreaks, the estimated mean 
$\mathcal{R}_0$ values were 1.5, 2.0, and 2.5. Estimate the final epidemic size 
for each outbreak.

```{r}
fixed_R0 <- 1.5
wrapper_objective_fn <- function(r_inf) objective_function(r_inf, fixed_R0)
res1 <- optimize(wrapper_objective_fn, interval = c(0, 1))

fixed_R0 <- 2
wrapper_objective_fn <- function(r_inf) objective_function(r_inf, fixed_R0)
res2 <- optimize(wrapper_objective_fn, interval = c(0, 1))

fixed_R0 <- 2.5
wrapper_objective_fn <- function(r_inf) objective_function(r_inf, fixed_R0)
res3 <- optimize(wrapper_objective_fn, interval = c(0, 1))

data.frame(R0    = c(1.5, 2.0, 2.5),
           r_inf = c(res1$minimum, res2$minimum, res3$minimum)) |> 
  gt() |> 
  fmt_number(
    columns = r_inf,
    decimals = 2)
```

# Numerically solving SIR models

We will introduce code to numerically integrate SIR models. We will use the 
deSolve package.

## Question 5

Create an SIR function that returns dS, dI, and dR for a given set of inputs: 
par_beta, par_gamma, and par_N (population size). Note that here we are working 
with the actual numbers of susceptible, infectious, and recovered individuals 
rather than their proportions.

```{r}
sir_model <- function(time, compartments, parameters) {
  with(as.list(c(compartments, parameters)), {
    
    dS <- -par_beta * S * I / par_N
    dI <-  par_beta * S * I / par_N - par_gamma * I
    dR <-  par_gamma * I
    
    list(c(dS, dI, dR))
  })
}
```

## Question 6

We can then use this code to numerically integrate an SIR model with fixed
parameters. Use an initial state of `S = 9999`, `I = 1`, `R = 0`, a 
`gamma of 0.5` and a `beta of 1`. Plot the resulting curves of S, I and R. 
Simulate for 60 days. 

```{r}
# Initial values
par_N        <- 10000
compartments <- c(S = 9999, I = 1, R = 0)

par_gamma <- 0.5

# Parameters
parameters <- c(
  par_beta = 1,   # transmission rate
  par_gamma = par_gamma,  # recovery rate
  par_N = par_N # population size
)

# Time points
stop_time <- 60
times <- seq(0, stop_time, by = 1/8)

# Solve ODE
out <- ode(
  y = compartments,
  times = times,
  func = sir_model,
  parms = parameters
) |> as.data.frame()
```

```{r}
tidy_out <- out |> 
  pivot_longer(-time, names_to = "Compartment") |> 
  mutate(beta = "1")

text_df <- data.frame(
  Compartment = c("S", "I", "R"),
  label = c("Susceptible", "Infectious", "Recovered"),
  xpos = c(40),
  ypos = c(2800, 600, 8500))

ggplot(tidy_out, aes(x = time, y = value)) +
  geom_line(aes(group = Compartment, colour = Compartment), linewidth = 2) +
  geom_text(data = text_df, aes(label = label, x = xpos, y = ypos, 
                                colour = Compartment), size = 7, hjust = 0) +
  scale_colour_manual(values = c("S" = "#5B573A",
                                 "I" = "#C96C3D",
                                 "R" = "#4C2218")) +
  labs(x = "Time (days)", y = "Value (people)") +
  theme_classic() +
  theme(axis.line = element_line(colour = "grey65"),
        legend.position = "none")
```


## Question 7

Run two more simulations. Change `par_beta` to `0.75` and `1.25`. All other
parameters and states should be the same as above. Plot the output
from these simulations on the same plot. Describe how these two epidemics
differ. Why would a different beta value result in a different 
epidemic?

```{r}
parameters["par_beta"] <- 0.75
out2 <- ode(
  y = compartments,
  times = times,
  func = sir_model,
  parms = parameters
) |> as.data.frame()

tidy_out2 <- out2 |> 
  pivot_longer(-time, names_to = "Compartment") |> 
  mutate(beta = "0.75")

parameters["par_beta"] <- 1.25

out3 <- ode(
  y = compartments,
  times = times,
  func = sir_model,
  parms = parameters
) |> as.data.frame()

tidy_out3 <- out3 |> 
  pivot_longer(-time, names_to = "Compartment") |> 
  mutate(beta = "1.25")

df <- bind_rows(tidy_out, tidy_out2, tidy_out3) |> 
  mutate(Compartment = factor(Compartment, levels = c("S", "I", "R")))

ggplot(df, aes(x = time, y = value)) +
  geom_line(aes(group = interaction(beta, Compartment), 
                colour = beta), linewidth = 2) +
  scale_colour_manual(values = c("1" = "#5B573A",
                                 "0.75" = "#C96C3D",
                                 "1.25" = "#4C2218")) +
  labs(x = "Time (days)", y = "Value (people)") +
  theme_classic() +
  facet_wrap(~Compartment, nrow = 1) +
  theme(axis.line = element_line(colour = "grey65"),
        legend.position = "bottom")
```

## Question 8

Estimate the final epidemic size of the three simulations above and compare
against those estimated in Question 4.

```{r}
r_inf_vals <- df |> filter(time == max(time),
             Compartment == "R") |> 
  mutate(r_inf = value / par_N) |> 
  pull(r_inf)

beta_vals <- c(1, 0.75, 1.25)

R0_vals <- beta_vals / par_gamma

data.frame(beta  = beta_vals,
           R0    = R0_vals,
           r_inf = r_inf_vals) |> 
  arrange(beta) |> 
  gt() |> 
    fmt_number(
    columns = r_inf,
    decimals = 2)
```

## Question 9

What's the largest value of `par_beta` for which the introduction of an 
infectious individuals does not result in an epidemic? Show your answer with a 
simulation. Simulate for 2000 days.

```{r}
parameters["par_beta"] <- 0.48

times <- seq(0, 2000, by = 1/8)

out4 <- ode(
  y = compartments,
  times = times,
  func = sir_model,
  parms = parameters
) |> as.data.frame()

tidy_out4 <- out4 |> 
  pivot_longer(-time, names_to = "Compartment") 

ggplot(tidy_out4, aes(x = time, y = value)) +
  geom_line(aes(group = Compartment, colour = Compartment), linewidth = 2) +
  scale_colour_manual(values = c("S" = "#5B573A",
                                 "I" = "#C96C3D",
                                 "R" = "#4C2218")) +
  labs(x = "Time (days)", y = "Value (people)") +
  theme_classic() +
  theme(axis.line = element_line(colour = "grey65"),
        legend.position = "none")
```














